---
title: "Creating plots in R using ggplot2 - part 11: linear regression plot"
author: 
  - Jodie Burchell
  - Mauricio Vargas Sepúlveda 帕夏
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, cache = FALSE, echo = FALSE, message = FALSE, warning = FALSE, tidy = FALSE}
knitr::opts_chunk$set(message = F, error = F, warning = F, comment = NA, fig.align = 'center', dpi = 100, fig.width=6, fig.height=5, tidy = F, cache.path = '.cache/', fig.path = '11_Linear_Regression_Plot/')
knitr::opts_knit$set(root.dir='11_Linear_Regression_Plot/')
```

This is the twelfth tutorial in a series on using `ggplot2` I am creating with [Mauricio Vargas Sepúlveda](http://pachamaltese.github.io/). In this tutorial we will demonstrate some of the many options the `ggplot2` package.

This post will be much more than just regression plot. Here we are getting, cleaning and processing financial data from [Quandl](https://www.quandl.com/). The goal is to estimate the CAPM model $R_i = R_f + \beta_i [R_m - R_f] + e_i$ where $R_i$ is the return of an asset, $R_f$ is the risk-free return (e.g. US Treasury Bonds), $R_m$ is the return of the market portfolio (e.g. NYSE) and $\beta_i$ is  a measure of risk relative to the market (e.g. $\beta_i = 1$ means that asset is exactly as risky as the market portfolio). More on the CAPM model can be read [here](http://people.stern.nyu.edu/ashapiro/courses/B01.231103/FFL09.pdf) but we will focus on plots.

The first thing to do is download and load in the data of the monthly price of Hang Seng Index (恒生指數) and Cheung Kong Holdings Hong Kong (长江实业集团香港) from 2015-03-01 to 2016-04-01. In order to work with Quandl without restrictions it is needed to create a free account and add `authcode = ...` to `Quandl` command.

```{r lr_load_in_data, cache=TRUE}
library(ggplot2)
library(gridExtra)
library(Quandl)
library(reshape)
library(ggfortify)

hsi.df <- Quandl("YAHOO/INDEX_HSI", start_date="2015-03-01", end_date="2016-04-01",
                 collapse="monthly", type = "raw", authcode = "XXX")

ckh.df <- Quandl("YAHOO/HK_0001", start_date="2015-03-01", 
                 end_date="2016-04-01", collapse="monthly", type = "raw", authcode="XXX")

saveRDS(hsi.df, "hsi.rds"); saveRDS(ckh.df,"ckh.rds")
```

In this tutorial, we will work towards creating the trend line and diagnostics plots below. We will take you from a basic density plot and explain all the customisations we add to the code step-by-step.

```{r lm_final, echo = FALSE, cache=TRUE}
hsi.df <- readRDS("hsi.rds")
colnames(hsi.df)[7] <- "Adjusted.Close"
hsi.df <- hsi.df[order(as.Date(hsi.df$Date)),]

hsi.Adjusted.Close <- hsi.df$Adjusted.Close
hsi.Return <- diff(hsi.Adjusted.Close)/hsi.Adjusted.Close[-length(hsi.Adjusted.Close)]
hsi.Return <- c(NA,hsi.Return)
hsi.df$Return <- hsi.Return
hsi.df <- na.omit(hsi.df)
hsi.Return <- hsi.df[,c("Date","Return")]

ckh.df <- readRDS("ckh.rds")
colnames(ckh.df)[7] <- "Adjusted.Close"
ckh.df <- ckh.df[order(as.Date(ckh.df$Date)),]
ckh.Adjusted.Close <- ckh.df$Adjusted.Close
ckh.Return <- diff(ckh.Adjusted.Close)/ckh.Adjusted.Close[-length(ckh.Adjusted.Close)]
ckh.Return <- c(NA,ckh.Return)
ckh.df <- na.omit(ckh.df)
ckh.df$Return <- ckh.Return
ckh.Return <- ckh.df[,c("Date","Return")]

hsi.ckh.returns <- merge(hsi.Return, ckh.Return, by='Date')
hsi.ckh.returns <- na.omit(hsi.ckh.returns)
colnames(hsi.ckh.returns) <- c("Date","hsi.Return","ckh.Return")

usa.risk.free <- 0.3/100
hsi.risk.premium <- 0.6/100

hsi.ckh.returns$hsi.Risk.free <- usa.risk.free + hsi.risk.premium
hsi.ckh.returns$hsi.Risk.premium <- hsi.ckh.returns$hsi.Return - hsi.ckh.returns$hsi.Risk.free

fit <- lm(ckh.Return ~ hsi.Risk.premium, data = hsi.ckh.returns) 

equation = function(x) {
  lm_coef <- list(a = round(coef(x)[1], digits = 2),
                  b = round(coef(x)[2], digits = 2),
                  r2 = round(summary(x)$r.squared, digits = 2));
  lm_eq <- substitute(italic(y) == a + b %.% italic(x)*","~~italic(R)^2~"="~r2,lm_coef)
  as.character(as.expression(lm_eq));                 
}

library(grid) 

fill <- "#4271AE"
lines <- "#1F3552"

p11 <- ggplot(hsi.ckh.returns, aes(x=hsi.Risk.premium, y=ckh.Return)) + geom_point(shape=1) + geom_smooth(method=lm, se=FALSE) +
      ggtitle("CKH regression line") +
      scale_x_continuous(name = "HSI risk premium") +
      scale_y_continuous(name = "CKH return") +
      annotate("rect", xmin = 0.00, xmax = 0.1, ymin = -0.056, ymax = -0.044, fill="white", colour="red") + 
      geom_text(x = 0.05, y = -0.05, label = equation(fit), parse = TRUE) + 
      theme(axis.line.x = element_line(size=.5, colour = "black"),
            axis.line.y = element_line(size=.5, colour = "black"),
            axis.text.x=element_text(colour="black", size = 9), 
            axis.text.y=element_text(colour="black", size = 9), 
            legend.position = "bottom", legend.position = "horizontal",
            panel.grid.major = element_line(colour = "#d3d3d3"), 
            panel.grid.minor = element_blank(), 
            panel.border = element_blank(), panel.background = element_blank(),
            plot.title = element_text(size = 14, family = "Tahoma", face = "bold"),
            text=element_text(family="Tahoma"))
p11

library(ggfortify)
autoplot(fit, label.size = 3) + theme(axis.line.x = element_line(size=.5, colour = "black"),
            axis.line.y = element_line(size=.5, colour = "black"),
            axis.text.x=element_text(colour="black", size = 9), 
            axis.text.y=element_text(colour="black", size = 9), 
            legend.position = "bottom", legend.position = "horizontal",
            panel.grid.major = element_line(colour = "#d3d3d3"), 
            panel.grid.minor = element_blank(), 
            panel.border = element_blank(), panel.background = element_blank(),
            plot.title = element_text(size = 14, family = "Tahoma", face = "bold"),
            text=element_text(family="Tahoma"))
```

Before calculating return as $R_i = \displaystyle \frac{P_t - P_{t-1}}{P_t}$ it is needed to order HSI and CKH data by dates and in decreasing order.

```{r lr_1, cache=TRUE}
hsi.df <- readRDS("hsi.rds")
colnames(hsi.df)[7] <- "Adjusted.Close"
hsi.df <- hsi.df[order(as.Date(hsi.df$Date)),]
```

With ordered dates it is possible to obtain the correct return for each month.

```{r lr_2, cache=TRUE}
hsi.Adjusted.Close <- hsi.df$Adjusted.Close
hsi.Return <- diff(hsi.Adjusted.Close)/hsi.Adjusted.Close[-length(hsi.Adjusted.Close)]
hsi.Return <- c(NA,hsi.Return)
hsi.df$Return <- hsi.Return
hsi.df <- na.omit(hsi.df)
hsi.Return <- hsi.df[,c("Date","Return")]

ckh.df <- readRDS("ckh.rds")
colnames(ckh.df)[7] <- "Adjusted.Close"
ckh.df <- ckh.df[order(as.Date(ckh.df$Date)),]
ckh.Adjusted.Close <- ckh.df$Adjusted.Close
ckh.Return <- diff(ckh.Adjusted.Close)/ckh.Adjusted.Close[-length(ckh.Adjusted.Close)]
ckh.Return <- c(NA,ckh.Return)
ckh.df <- na.omit(ckh.df)
ckh.df$Return <- ckh.Return
ckh.Return <- ckh.df[,c("Date","Return")]
```

The returns can be arranged in one data frame before doing plots and regression.

```{r lr_3, cache=TRUE}
hsi.ckh.returns <- merge(hsi.Return, ckh.Return, by='Date')
hsi.ckh.returns <- na.omit(hsi.ckh.returns)
colnames(hsi.ckh.returns) <- c("Date","hsi.Return","ckh.Return")
```

Using [Damodaran](http://pages.stern.nyu.edu/~adamodar/New_Home_Page/datafile/ctryprem.html) and [Bloomberg](http://www.bloomberg.com/markets/rates-bonds/government-bonds/us) data we can work with an estimate of HSI risk premium over risk-free rate.

```{r lr_4, cache=TRUE}
usa.risk.free <- 0.3/100
hsi.risk.premium <- 0.6/100
```

## Trend line plot

### Basic trend line plot

Now we can fit a linear regression. One interesting thing is that in CAPM context the regression line slope can be calculated as $\beta_i = \displaystyle \frac{\sigma_{i,m}}{\sigma_m^2}$.

```{r lr_5, cache=TRUE}
hsi.ckh.returns$hsi.Risk.free <- usa.risk.free + hsi.risk.premium
hsi.ckh.returns$hsi.Risk.premium <- hsi.ckh.returns$hsi.Return - hsi.ckh.returns$hsi.Risk.free
hsi.Return.vector <- as.vector(hsi.ckh.returns$hsi.Return)
ckh.Return.vector <- as.vector(hsi.ckh.returns$ckh.Return)
cov.hsi.ckh <- cov(ckh.Return.vector, hsi.Return.vector)
var.hk <- var(hsi.Return.vector)
capm_beta = cov.hsi.ckh/var.hk

fit <- lm(ckh.Return ~ hsi.Risk.premium, data = hsi.ckh.returns) 
summary(fit)
```

Up to this point we have all what is required to plot regressions. We will start with a basic regression plot.

```{r lr_6, cache=TRUE}
p11 <- ggplot(hsi.ckh.returns, aes(x=hsi.Risk.premium, y=ckh.Return)) + geom_point(shape=1) + geom_smooth(method=lm) 
p11
```

`geom_point` can be customized, for example, not to include the confidence region

```{r lr_7, cache=TRUE}
p11 <- ggplot(hsi.ckh.returns, aes(x=hsi.Risk.premium, y=ckh.Return)) + geom_point(shape=1) + geom_smooth(method=lm, se=FALSE) 
p11
```

Before continuing it is a good idea to fix the axis labels and add a title.

### Customising axis labels

```{r lr_8, cache=TRUE}
p11 <- p11 + scale_x_continuous(name = "HSI risk premium") + 
      scale_y_continuous(name = "CKH return")
p11
```

### Adding a title

```{r lr_9, cache=TRUE}
p11 <- p11 + ggtitle("CKH regression line")
p11
```

### Including regression coefficients

It would be interesting to show $R^2$ and regression coefficients within the plot.

```{r lr_10, cache=TRUE}
p11 <- p11 + annotate("text", x=0.1, y=-0.05, label = "R^2=0.78") + 
      annotate("text", x=0.1, y=-0.06, label = "alpha=0.00") + 
      annotate("text", x=0.1, y=-0.07, label = "beta=0.67")
p11
```

Another option would be to add greek letters and exponents.

```{r lr_11, cache=TRUE}
p11 <- ggplot(hsi.ckh.returns, aes(x=hsi.Risk.premium, y=ckh.Return)) + geom_point(shape=1) + 
      geom_smooth(method=lm, se=FALSE) + ggtitle("CKH regression line") +
      scale_x_continuous(name = "HSI risk premium") +
      scale_y_continuous(name = "CKH return") + 
      annotate("text", x=0.1, y=-0.05, label = "R^2 == 0.78", parse=T) + 
      annotate("text", x=0.1, y=-0.06, label = "alpha == 0.00", parse=T) + 
      annotate("text", x=0.1, y=-0.07, label = "beta == 0.67", parse=T)
p11
```

To make the coefficients more visible we will add some elements to increase visibility

```{r lr_12, cache=TRUE}
p11 <- ggplot(hsi.ckh.returns, aes(x=hsi.Risk.premium, y=ckh.Return)) + 
      geom_point(shape=1) + geom_smooth(method=lm, se=FALSE) +
      ggtitle("CKH regression line") +
      scale_x_continuous(name = "HSI risk premium") +
      scale_y_continuous(name = "CKH return") +
      annotate("rect", xmin = 0.075, xmax = 0.125, ymin = -0.075, ymax = -0.045, fill="white", colour="red") + 
      annotate("text", x=0.1, y=-0.05, label = "R^2 == 0.78", parse=T) + annotate("text", x=0.1, y=-0.06, label = "alpha == 0.00", parse=T) + 
      annotate("text", x=0.1, y=-0.07, label = "beta == 0.67", parse=T)
p11
```

Another customization could be to show the trend line using rounded digits (or even significative digits) from regression coefficients. This requires us to write a function and is not as easy to obtain as the last plot. 

```{r lr_13, cache=TRUE}
equation = function(x) {
  lm_coef <- list(a = round(coef(x)[1], digits = 2),
                  b = round(coef(x)[2], digits = 2),
                  r2 = round(summary(x)$r.squared, digits = 2));
  lm_eq <- substitute(italic(y) == a + b %.% italic(x)*","~~italic(R)^2~"="~r2,lm_coef)
  as.character(as.expression(lm_eq));                 
}

p11 <- ggplot(hsi.ckh.returns, aes(x=hsi.Risk.premium, y=ckh.Return)) + geom_point(shape=1) + geom_smooth(method=lm, se=FALSE) +
      ggtitle("CKH regression line") +
      scale_x_continuous(name = "HSI risk premium") +
      scale_y_continuous(name = "CKH return") +
      annotate("rect", xmin = 0.00, xmax = 0.1, ymin = -0.056, ymax = -0.044, fill="white", colour="red") + 
      geom_text(x = 0.05, y = -0.05, label = equation(fit), parse = TRUE)
p11
```

### Using the white theme

```{r lr_14, cache=TRUE}
p11 <- p11 + theme_bw()
p11
```

### Creating an XKCD style chart

Of course, you may want to create your own themes as well. `ggplot2` allows for a very high degree of customisation, including allowing you to use imported fonts. Below is an example of a theme Mauricio was able to create which mimics the visual style of [XKCD](http://xkcd.com/). In order to create this chart, you first need to import the XKCD font, install it on your machine and load it into R using the `extrafont` package.
These instructions are taken from [here](https://www.google.com.au/url?sa=t&rct=j&q=&esrc=s&source=web&cd=1&ved=0ahUKEwiWzafchdPJAhVBpJQKHe_LDT8QFggbMAA&url=https%3A%2F%2Fcran.r-project.org%2Fweb%2Fpackages%2Fxkcd%2Fvignettes%2Fxkcd-intro.pdf&usg=AFQjCNE-KciGY14e-Q1buYIVmTFC0ht__Q&sig2=DZUwkvIHwfNWtTtkcz94jg):

```{r lr_importing_font_xkcd, eval=FALSE}
library(extrafont)

download.file("http://simonsoftware.se/other/xkcd.ttf", 
              dest="xkcd.ttf", mode="wb")
system("mkdir ~/.fonts")
system("cp xkcd.ttf  ~/.fonts")
font_import(paths = "~/.fonts", pattern="[X/x]kcd")
fonts()
loadfonts()
```

You can then create your graph:

```{r lr_15, cache=TRUE}
p11 <- ggplot(hsi.ckh.returns, aes(x=hsi.Risk.premium, y=ckh.Return)) + geom_point(shape=1) + geom_smooth(method=lm, se=FALSE) +
      ggtitle("CKH regression line") +
      scale_x_continuous(name = "HSI risk premium") +
      scale_y_continuous(name = "CKH return") +
      annotate("rect", xmin = 0.00, xmax = 0.1, ymin = -0.056, ymax = -0.044, fill="white", colour="red") + 
      geom_text(x = 0.05, y = -0.05, label = equation(fit), parse = TRUE) + 
      theme(axis.line.x = element_line(size=.5, colour = "black"),
            axis.line.y = element_line(size=.5, colour = "black"), 
            axis.text.x=element_text(colour="black", size = 9), 
            axis.text.y=element_text(colour="black", size = 9),
            panel.grid.major = element_line(colour = "#d3d3d3"), 
            panel.grid.minor = element_blank(), 
            panel.border = element_blank(), panel.background = element_blank(),
            plot.title = element_text(family = "xkcd-Regular"),
            text=element_text(family="xkcd-Regular"))
p11
```

### Using 'The Economist' theme 

There are a wider range of pre-built themes available as part of the `ggthemes` package (more information on these [here](https://cran.r-project.org/web/packages/ggthemes/vignettes/ggthemes.html)). Below we've applied `theme_economist()`, which approximates graphs in the Economist magazine.

```{r lr_16, cache=TRUE}
library(ggthemes)
library(grid)

fill <- "#4271AE"
line <- "#1F3552"

p11 <- ggplot(hsi.ckh.returns, aes(x=hsi.Risk.premium, y=ckh.Return)) + geom_point(shape=1) + geom_smooth(method=lm, se=FALSE) +
      ggtitle("CKH regression line") +
      scale_x_continuous(name = "HSI risk premium") +
      scale_y_continuous(name = "CKH return") +
      annotate("rect", xmin = 0.00, xmax = 0.1, ymin = -0.056, ymax = -0.044, fill="white", colour="red") + 
      geom_text(x = 0.05, y = -0.05, label = equation(fit), parse = TRUE) + 
      theme_economist() +
      theme(axis.line.x = element_line(size=.5, colour = "black"),
            axis.line.y = element_line(size=.5, colour = "black"), 
            axis.text.x=element_text(colour="black", size = 9), 
            axis.text.y=element_text(colour="black", size = 9),
            panel.grid.major = element_line(colour = "#d3d3d3"), 
            panel.grid.minor = element_blank(), 
            panel.border = element_blank(), panel.background = element_blank(),
            plot.title = element_text(family = "OfficinaSanITC-Book"),
            text=element_text(family="OfficinaSanITC-Book"))
p11
```

### Creating your own theme

As before, you can modify your plots a lot as `ggplot2` allows many customisations. Here is a custom plot where we have modified the axes, background and font.

```{r lr_17, cache=TRUE}
library(grid) 

fill <- "#4271AE"
lines <- "#1F3552"

p11 <- ggplot(hsi.ckh.returns, aes(x=hsi.Risk.premium, y=ckh.Return)) + geom_point(shape=1) + geom_smooth(method=lm, se=FALSE) +
      ggtitle("CKH regression line") +
      scale_x_continuous(name = "HSI risk premium") +
      scale_y_continuous(name = "CKH return") +
      annotate("rect", xmin = 0.00, xmax = 0.1, ymin = -0.056, ymax = -0.044, fill="white", colour="red") + 
      geom_text(x = 0.05, y = -0.05, label = equation(fit), parse = TRUE) + 
      theme(axis.line.x = element_line(size=.5, colour = "black"),
            axis.line.y = element_line(size=.5, colour = "black"),
            axis.text.x=element_text(colour="black", size = 9), 
            axis.text.y=element_text(colour="black", size = 9), 
            legend.position = "bottom", legend.position = "horizontal",
            panel.grid.major = element_line(colour = "#d3d3d3"), 
            panel.grid.minor = element_blank(), 
            panel.border = element_blank(), panel.background = element_blank(),
            plot.title = element_text(size = 14, family = "Tahoma", face = "bold"),
            text=element_text(family="Tahoma"))
p11
```

## Regression diagnostics plots

### Basic diagnostics plots

`ggfortify` lets `ggplot2` know how to interpret `lm` objects.

```{r lr_18, cache=TRUE}
# install.packages("ggfortify")
library(ggfortify)
autoplot(fit, label.size = 3)
```

### Using the white theme

```{r lr_19, cache=TRUE}
autoplot(fit, label.size = 3) + theme_bw()
```

### Creating an XKCD style chart

```{r lr_20, cache=TRUE}
autoplot(fit, label.size = 3) + theme(axis.line.x = element_line(size=.5, colour = "black"),
            axis.line.y = element_line(size=.5, colour = "black"), 
            axis.text.x=element_text(colour="black", size = 9), 
            axis.text.y=element_text(colour="black", size = 9),
            panel.grid.major = element_line(colour = "#d3d3d3"), 
            panel.grid.minor = element_blank(), 
            panel.border = element_blank(), panel.background = element_blank(),
            plot.title = element_text(family = "xkcd-Regular"),
            text=element_text(family="xkcd-Regular"))
```

### Using 'The Economist' theme 

```{r lr_21, cache=TRUE}
autoplot(fit, label.size = 3) + theme_economist() +
      theme(axis.line.x = element_line(size=.5, colour = "black"),
            axis.line.y = element_line(size=.5, colour = "black"), 
            axis.text.x=element_text(colour="black", size = 9), 
            axis.text.y=element_text(colour="black", size = 9),
            panel.grid.major = element_line(colour = "#d3d3d3"), 
            panel.grid.minor = element_blank(), 
            panel.border = element_blank(), panel.background = element_blank(),
            plot.title = element_text(family = "OfficinaSanITC-Book"),
            text=element_text(family="OfficinaSanITC-Book"))
```

### Creating your own theme

```{r lr_22, cache=TRUE}
autoplot(fit, label.size = 3) + theme(axis.line.x = element_line(size=.5, colour = "black"),
            axis.line.y = element_line(size=.5, colour = "black"),
            axis.text.x=element_text(colour="black", size = 9), 
            axis.text.y=element_text(colour="black", size = 9), 
            legend.position = "bottom", legend.position = "horizontal",
            panel.grid.major = element_line(colour = "#d3d3d3"), 
            panel.grid.minor = element_blank(), 
            panel.border = element_blank(), panel.background = element_blank(),
            plot.title = element_text(size = 14, family = "Tahoma", face = "bold"),
            text=element_text(family="Tahoma"))
```
