---
title: "Creating plots in R using ggplot2 - part 11: linear regression plot"
author: 
  - Jodie Burchell
  - Mauricio Vargas Sepúlveda 帕夏
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, cache = FALSE, echo = FALSE, message = FALSE, warning = FALSE, tidy = FALSE}
knitr::opts_chunk$set(message = F, error = F, warning = F, comment = NA, fig.align = 'center', dpi = 100, fig.width=6, fig.height=5, tidy = F, cache.path = '.cache/', fig.path = '11_Linear_Regression_Plot/')
knitr::opts_knit$set(root.dir='11_Linear_Regression_Plot/')
```

This is the twelfth tutorial in a series on using `ggplot2` I am creating with [Mauricio Vargas Sepúlveda](http://pachamaltese.github.io/). In this tutorial we will demonstrate some of the many options the `ggplot2` package.

This post will be much more than just regression plot. Here we are getting, cleaning and processing financial data from [Quandl](https://www.quandl.com/). The goal is to estimate the CAPM model $R_i = R_f + \beta_i [R_m - R_f] + e_i$ where $R_i$ is the return of an asset, $R_f$ is the risk-free return (e.g. US Treasury Bonds), $R_m$ is the return of the market portfolio (e.g. NYSE) and $\beta_i$ is  a measure of risk relative to the market (e.g. $\beta_i = 1$ means that asset is exactly as risky as the market portfolio). More on the CAPM model can be read [here](http://people.stern.nyu.edu/ashapiro/courses/B01.231103/FFL09.pdf) but we will focus on plots.

The first thing to do is download and load in the data of the monthly price of Hang Seng Index (恒生指數) and Cheung Kong Holdings Hong Kong (长江实业集团香港) from 2015-03-01 to 2016-04-01. In order to work with Quandl without restrictions it is needed to create a free account and add `authcode = ...` to `Quandl` command.

```{r lr_load_in_data}
library(ggplot2);library(gridExtra);library(Quandl);library(reshape)
```

```{r lr_2, eval=FALSE}
hsi.df <- Quandl("YAHOO/INDEX_HSI", start_date="2015-03-01", end_date="2016-04-01", 
                collapse="monthly", type = "raw", authcode = "XXX")

ckh.df <- Quandl("YAHOO/HK_0001", start_date="2015-03-01", 
                 end_date="2016-04-01", collapse="monthly", type = "raw", authcode="XXX")

saveRDS(hsi.df, "hsi.rds"); saveRDS(ckh.df,"ckh.rds")
```

Before calculating return as $R_i = \displaystyle \frac{P_t - P_{t-1}}{P_t}$ it is needed to order HSI and CKH data by dates in decreasing order.

```{r lr_3, cache=TRUE}
hsi.df <- readRDS("hsi.rds")
colnames(hsi.df)[7] <- "Adjusted.Close"
hsi.df <- hsi.df[order(as.Date(hsi.df$Date)),]
```

With ordered dates it is possible to obtain the correct return for each month.

```{r lr_4, cache=TRUE}
hsi.Adjusted.Close <- hsi.df$Adjusted.Close
hsi.Return <- diff(hsi.Adjusted.Close)/hsi.Adjusted.Close[-length(hsi.Adjusted.Close)]
hsi.Return <- c(NA,hsi.Return)
hsi.df$Return <- hsi.Return
hsi.df <- na.omit(hsi.df)
hsi.Return <- hsi.df[,c("Date","Return")]

ckh.df <- readRDS("ckh.rds")
colnames(ckh.df)[7] <- "Adjusted.Close"
ckh.df <- ckh.df[order(as.Date(ckh.df$Date)),]
ckh.Adjusted.Close <- ckh.df$Adjusted.Close
ckh.Return <- diff(ckh.Adjusted.Close)/ckh.Adjusted.Close[-length(ckh.Adjusted.Close)]
ckh.Return <- c(NA,ckh.Return)
ckh.df <- na.omit(ckh.df)
ckh.df$Return <- ckh.Return
ckh.Return <- ckh.df[,c("Date","Return")]
```

With correct returns and ordered by date prices it is possible to plot HSI (恒生指數) price and return.

```{r lr_5, cache=TRUE}
p1 <- ggplot(hsi.df)
p1 <- p1 + geom_line(aes(x = Date, y = Adjusted.Close), col="darkblue")
p1 <- p1 + labs(title = "Hang Seng Index", x = "Date", y = "Nominal Price (HKD)")

p2 <- ggplot(hsi.df)
p2 <- p2 + geom_line(aes(x = Date, y = Return), col="darkred")
p2 <- p2 + labs(title = "", x = "Date", y = "Monthly Return (%)")

grid.arrange(p1, p2, nrow=2)

p3 <- ggplot(ckh.df)
p3 <- p3 + geom_line(aes(x = Date, y = Adjusted.Close), col="darkblue")
p3 <- p3 + labs(title = "Cheung Kong Holdings Hong Kong", x = "Date", y = "Nominal Price (HKD)")

p4 <- ggplot(ckh.df)
p4 <- p4 + geom_line(aes(x = Date, y = Return), col="darkred")
p4 <- p4 + labs(title = "", x = "Date", y = "Monthly Return (%)")

grid.arrange(p3, p4, nrow=2)
```

Before doing regression we will plot the return distribution and test normality in return distribution. In this case the test must be Shapiro-Wilk as Kolmogorov-Smirov requieres a continuous distribution.

```{r lr_6, cache=TRUE}
p5 <- ggplot(hsi.df, aes(x = Return)) + geom_density()
p5

p6 <- ggplot(ckh.df, aes(x = Return)) + geom_density()
p6

shapiro.test(hsi.df$Return)
shapiro.test(ckh.df$Return)
```

The returns can be arranged in one data frame before doing plots and regression.

```{r lr_7, cache=TRUE}
hsi.ckh.returns <- merge(hsi.Return, ckh.Return, by='Date')
hsi.ckh.returns <- na.omit(hsi.ckh.returns)
colnames(hsi.ckh.returns) <- c("Date","hsi.Return","ckh.Return")
```

With the previous arrangements, now it is possible to show the returns in one plot.

```{r lr_8, cache=TRUE, fig.width=8, fig.height=6}
hsi.ckh.returns.melt <- melt(hsi.ckh.returns, id="Date")  # convert to long format
p7 <- ggplot(data=hsi.ckh.returns.melt, aes(x=Date, y=value, colour=variable))
p7 <- p7 + geom_point() + geom_line()
p7 <- p7 + scale_color_manual("Variable",labels = c("HSI", "CKH"), values = c("royalblue", "red"))
p7 <- p7 + labs(title="Index Comparison", x="Date", y="Return (%)")
p7
```

Using [Damodaran](http://pages.stern.nyu.edu/~adamodar/New_Home_Page/datafile/ctryprem.html) and [Bloomberg](http://www.bloomberg.com/markets/rates-bonds/government-bonds/us) data we can work with an estimate of HSI risk premium over risk-free rate.

```{r lr_9, cache=TRUE}
usa.risk.free <- 0.3/100
hsi.risk.premium <- 0.6/100

hsi.ckh.returns$hsi.Risk.free <- usa.risk.free + hsi.risk.premium
hsi.ckh.returns$hsi.Risk.premium <- hsi.ckh.returns$hsi.Return - hsi.ckh.returns$hsi.Risk.free
```

Now we can fit a lineal regression. One interesting thing is that in CAPM context the regression line slope can be calculated as $\beta_i = \displaystyle \frac{\sigma_{i,m}}{\sigma_m^2}$.

```{r lr_10, cache=TRUE}
hsi.Return.vector <- as.vector(hsi.ckh.returns$hsi.Return)
ckh.Return.vector <- as.vector(hsi.ckh.returns$ckh.Return)
cov.hsi.ckh <- cov(ckh.Return.vector, hsi.Return.vector)
var.hk <- var(hsi.Return.vector)
cov.hsi.ckh/var.hk

fit <- lm(ckh.Return ~ hsi.Risk.premium, data = hsi.ckh.returns) 
summary(fit)
```

Up to this point we have all what is required to plot regressions. We will start with a basic regression plot.

```{r lr_11, cache=TRUE}
p8 <- ggplot(hsi.ckh.returns, aes(x=hsi.Risk.premium, y=ckh.Return)) + geom_point(shape=1) + geom_smooth(method=lm) 
p8
```

`geom_point` can be customized, for example, not to include the confidence region

```{r lr_12, cache=TRUE}
p8 <- ggplot(hsi.ckh.returns, aes(x=hsi.Risk.premium, y=ckh.Return)) + geom_point(shape=1) + geom_smooth(method=lm, se=FALSE) 
p8
```

Before continuing it is a good idea to fix the axis names and add a title

```{r lr_13, cache=TRUE}
p8 <- ggplot(hsi.ckh.returns, aes(x=hsi.Risk.premium, y=ckh.Return)) + geom_point(shape=1) + geom_smooth(method=lm, se=FALSE) + ggtitle("CKH regression line") +
  scale_x_continuous(name = "HSI risk premium") +
  scale_y_continuous(name = "CKH return")
p8
```

It would be interesting to show $R^2$ and regression coefficients within the plot. This can be more trickier but it is possible.

```{r lr_14, cache=TRUE}
p8 <- ggplot(hsi.ckh.returns, aes(x=hsi.Risk.premium, y=ckh.Return)) + geom_point(shape=1) + geom_smooth(method=lm, se=FALSE) + ggtitle("CKH regression line") +
  scale_x_continuous(name = "HSI risk premium") +
  scale_y_continuous(name = "CKH return") + annotate("text", x=0.1, y=-0.05, label = "R^2=0.78") + annotate("text", x=0.1, y=-0.06, label = "alpha=0.00") + annotate("text", x=0.1, y=-0.07, label = "beta=0.67")
p8
```

Now we can add the greek letters and exponents.

```{r lr_15, cache=TRUE}
p8 <- ggplot(hsi.ckh.returns, aes(x=hsi.Risk.premium, y=ckh.Return)) + geom_point(shape=1) + geom_smooth(method=lm, se=FALSE) + ggtitle("CKH regression line") +
  scale_x_continuous(name = "HSI risk premium") +
  scale_y_continuous(name = "CKH return") + annotate("text", x=0.1, y=-0.05, label = "R^2 == 0.78", parse=T) + annotate("text", x=0.1, y=-0.06, label = "alpha == 0.00", parse=T) + annotate("text", x=0.1, y=-0.07, label = "beta == 0.67", parse=T)
p8
```

To make the coefficients more visible we will add some elements to increase visibility

```{r lr_16, cache=TRUE}
p8 <- ggplot(hsi.ckh.returns, aes(x=hsi.Risk.premium, y=ckh.Return)) + geom_point(shape=1) + geom_smooth(method=lm, se=FALSE) 
p8 <- p8 + 
  ggtitle("CKH regression line") +
  scale_x_continuous(name = "HSI risk premium") +
  scale_y_continuous(name = "CKH return") +
  annotate("rect", xmin = 0.075, xmax = 0.125, ymin = -0.075, ymax = -0.045, fill="white", colour="red") + 
  annotate("text", x=0.1, y=-0.05, label = "R^2 == 0.78", parse=T) + annotate("text", x=0.1, y=-0.06, label = "alpha == 0.00", parse=T) + 
  annotate("text", x=0.1, y=-0.07, label = "beta == 0.67", parse=T)
p8
```