---
title: "Creating basic graphs in R"
output: html_document
---

This tutorial will demonstrate how to create graphs in ggplot2. It will demonstrate the variety of graphs you can make without further customisation.

## List of graphs:
* Scatterplot
* Weighted scatterplot
* Linear regression plot
* Lowess curve plot
* Line graph
* Area graph
* Bar graph OK
* Stacked bar graph OK
* Box plot
* Histogram
* Density plot
* Function curve

## Stacked bar graph

```{r, cache=TRUE}
library(ggplot2)

barcharts.data <- read.csv("barcharts.csv")
p9 <- ggplot() + geom_bar(aes(y = export, x = year, fill = product), data = barcharts.data, stat="identity")
p9
```

```{r, cache=TRUE}
p9 <- p9 + geom_text(data=barcharts.data, aes(x = year, y = export, label = export, size=4))
p9
```

```{r, cache=TRUE}
p9 <- ggplot() + 
  geom_bar(aes(y = export, x = year, fill = product), data = barcharts.data, stat="identity") +
  geom_text(data=barcharts.data, aes(x = year, y = export, label = export, size=4), show_guide = F)
p9
```

```{r, cache=TRUE}
library(plyr)
barcharts.data <- as.data.frame(barcharts.data)
barcharts.data$product <- factor(barcharts.data$product, levels = c("copper","others"), labels = c("Copper","Pulp wood, Fruit, Salmon & Others"))
barcharts.data <- ddply(barcharts.data, .(year), transform, pos = cumsum(export) - (0.5 * export))

p9 <- ggplot() + geom_bar(aes(y = export, x = year, fill = product), data = barcharts.data, stat="identity")
p9 <- p9 + geom_text(data=barcharts.data, aes(x = year, y = pos, label = export, size=4), show_guide = F)
p9
```

```{r, cache=TRUE}
p9 <- p9 + theme(legend.position="bottom", legend.direction="horizontal", legend.title = element_blank())
p9
```

```{r, cache=TRUE}
library(scales)
p9 <- p9 + scale_x_continuous(breaks=seq(2006,2014,1))
p9
```

```{r, cache=TRUE}
p9 <- p9 + labs(x="Year", y="USD million") + ggtitle("Composition of Exports to China ($)") 
p9
```

```{r, cache=TRUE}
fill <- c("#56B4E9", "#F0E442")
p9 <- p9 + scale_fill_manual(values=fill)
p9
```

```{r, cache=TRUE}
p9 <- ggplot() + theme_bw() +
  geom_bar(aes(y = export, x = year, fill = product), data = barcharts.data, stat="identity") + 
  geom_text(data=barcharts.data, aes(x = year, y = pos, label = export, size=4), show_guide = F) + 
  theme(legend.position="bottom", legend.direction="horizontal", legend.title = element_blank()) + 
  scale_x_continuous(breaks=seq(2006,2014,1)) + 
  labs(x="Year", y="USD million") + 
  ggtitle("Composition of Exports to China ($)") + 
  scale_fill_manual(values=fill)
p9
```

```{r, cache=TRUE}
p9 <- ggplot() + 
  geom_bar(aes(y = export, x = year, fill = product), data = barcharts.data, stat="identity") + 
  geom_text(data=barcharts.data, aes(x = year, y = pos, label = export), colour="black", family="AdobeHeitiStd-Regular", size = 4, show_guide = F) + 
  theme(legend.position="bottom", legend.direction="horizontal", legend.title = element_blank()) + 
  scale_x_continuous(breaks=seq(2006,2014,1)) + 
  labs(x="Year", y="USD million") + 
  ggtitle("Composition of Exports to China ($)") + 
  scale_fill_manual(values=fill) + 
  theme(axis.line = element_line(size=1)) + 
  theme(axis.line = element_line(colour = "black"), panel.grid.major = element_line(colour = "#d3d3d3"), panel.grid.minor = element_blank(), panel.border = element_blank(), panel.background = element_blank()) + 
  theme(plot.title=element_text(family="AdobeHeitiStd-Regular"), text=element_text(family="AdobeHeitiStd-Regular"), axis.text.x=element_text(colour="black"), axis.text.y=element_text(colour="black")) 
p9
```

```{r, cache=TRUE}
p9 <- ggplot() + 
  geom_bar(aes(y = export, x = year, fill = product), data = barcharts.data, stat="identity") + 
  geom_text(data=barcharts.data, aes(x = year, y = pos, label = export), colour="black", family="xkcd-Regular", size = 4, show_guide = F) + 
  theme(legend.position="bottom", legend.direction="horizontal", legend.title = element_blank()) + 
  scale_x_continuous(breaks=seq(2006,2014,1)) + 
  labs(x="Year", y="USD million") + 
  ggtitle("Composition of Exports to China ($)") + 
  scale_fill_manual(values=fill) + 
  theme(axis.line = element_line(size=1)) + 
  theme(axis.line = element_line(colour = "black"), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.border = element_blank(), panel.background = element_blank()) + 
  theme(plot.title=element_text(family="xkcd-Regular"), text=element_text(family="xkcd-Regular"), axis.text.x=element_text(colour="black"), axis.text.y=element_text(colour="black")) 
p9
```

```{r, cache=TRUE}
# install.packages('ggthemes', dependencies = TRUE)
library(ggthemes)
library(extrafont)

fill <- c("#5b92a7", "#aaa9ac")
p9 <- ggplot() + theme_economist() + scale_fill_economist() +   theme(plot.title=element_text(family="OfficinaSanITC-Book"), text=element_text(family="OfficinaSanITC-Book")) +
  geom_bar(aes(y = export, x = year, fill = product), data = barcharts.data, stat="identity") + 
  geom_text(data=barcharts.data, aes(x = year, y = pos, label = export), colour="white", size = 4, family = "OfficinaSanITC-Book", show_guide = F) + 
  theme(legend.position="bottom", legend.direction="horizontal", legend.title = element_blank()) + 
  scale_x_continuous(breaks=seq(2006,2014,1)) + 
  labs(x="Year", y="USD million") + 
  ggtitle("Composition of Exports to China ($)")
p9
```

## Stacked bar graph

```{r, cache=TRUE}
library(ggplot2)

barcharts.data <- read.csv("barcharts.csv")
p10 <- ggplot() + geom_bar(aes(y = percentage, x = year, fill = product), data = barcharts.data, stat="identity")
p10
```

```{r, cache=TRUE}
p10 <- p10 + geom_text(data=barcharts.data, aes(x = year, y = percentage, label = paste0(percentage,"%")), show_guide = F)
p10
```

```{r, cache=TRUE}
library(plyr)
barcharts.data <- as.data.frame(barcharts.data)
barcharts.data$product <- factor(barcharts.data$product, levels = c("copper","others"), labels = c("Copper","Pulp wood, Fruit, Salmon & Others"))
barcharts.data <- ddply(barcharts.data, .(year), transform, pos = cumsum(percentage) - (0.5 * percentage))

p10 <- ggplot() + geom_bar(aes(y = percentage, x = year, fill = product), data = barcharts.data, stat="identity")
p10 <- p10 + geom_text(data=barcharts.data, aes(x = year, y = pos, label = paste0(percentage,"%")), show_guide = F)
p10
```

```{r, cache=TRUE}
p10 <- p10 + theme(legend.position="bottom", legend.direction="horizontal", legend.title = element_blank())
p10
```

```{r, cache=TRUE}
library(scales)
p10 <- p10 + scale_x_continuous(breaks=seq(2006,2014,1)) + scale_y_continuous(labels = dollar_format(suffix = "%", prefix = ""))
p10
```

```{r, cache=TRUE}
p10 <- p10 + labs(x="Year", y="Percentage") + ggtitle("Composition of Exports to China (%)") 
p10
```

```{r, cache=TRUE}
fill <- c("#56B4E9", "#F0E442")
p10 <- p10 + scale_fill_manual(values=fill)
p10
```

```{r, cache=TRUE}
p10 <- ggplot() + 
  geom_bar(aes(y = percentage, x = year, fill = product), data = barcharts.data, stat="identity") + 
  geom_text(data=barcharts.data, aes(x = year, y = pos, label = paste0(percentage,"%")), colour="black", family="AdobeHeitiStd-Regular", size = 5, show_guide = F) + 
  theme(legend.position="bottom", legend.direction="horizontal", legend.title = element_blank()) + 
  scale_x_continuous(breaks=seq(2006,2014,1)) + 
  scale_y_continuous(labels = dollar_format(suffix = "%", prefix = "")) + 
  labs(x="Year", y="Percentage") + 
  ggtitle("Composition of Exports to China (%)") + 
  scale_fill_manual(values=fill) + 
  theme(axis.line = element_line(size=1)) + 
  theme(axis.line = element_line(colour = "black"), panel.grid.major = element_line(colour = "#d3d3d3"), panel.grid.minor = element_blank(), panel.border = element_blank(), panel.background = element_blank()) + 
  theme(plot.title=element_text(family="AdobeHeitiStd-Regular"), text=element_text(family="AdobeHeitiStd-Regular"), axis.text.x=element_text(colour="black"), axis.text.y=element_text(colour="black")) 
p10
```

```{r, cache=TRUE}
p10 <- ggplot() + 
  geom_bar(aes(y = percentage, x = year, fill = product), data = barcharts.data, stat="identity") + 
  geom_text(data=barcharts.data, aes(x = year, y = pos, label = paste0(percentage,"%")), colour="black", family="xkcd-Regular", size = 5, show_guide = F) + 
  theme(legend.position="bottom", legend.direction="horizontal", legend.title = element_blank()) + 
  scale_x_continuous(breaks=seq(2006,2014,1)) + 
  scale_y_continuous(labels = dollar_format(suffix = "%", prefix = "")) + 
  labs(x="Year", y="Percentage") + 
  ggtitle("Composition of Exports to China (%)") + 
  scale_fill_manual(values=fill) + 
  theme(axis.line = element_line(size=1)) + 
  theme(axis.line = element_line(colour = "black"), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.border = element_blank(), panel.background = element_blank()) + 
  theme(plot.title=element_text(family="xkcd-Regular"), text=element_text(family="xkcd-Regular"), axis.text.x=element_text(colour="black"), axis.text.y=element_text(colour="black")) 
p10
```






