---
title: "Creating plots in R using ggplot2 - Parts 1-10"
author:
- Jodie Burchell
- Mauricio Vargas Sepúlveda
date: '`r Sys.Date()`'
output:
  pdf_document:
    keep_tex: yes
    toc: yes
---

```{r setup, cache = FALSE, echo = FALSE, message = FALSE, warning = FALSE, tidy = FALSE}
knitr::opts_chunk$set(message = F, error = F, warning = F, comment = NA, fig.align = 'center', dpi = 96, fig.width=8, fig.height=6, tidy = F, cache.path = '.cache/', fig.path = '0_all_posts_pdf/', dev = 'quartz_pdf', dev.args=list(pointsize=12))
```

I recently teamed up with [Mauricio Vargas Sepúlveda](http://pachamaltese.github.io/) to create some graphing tutorials in R. In the coming weeks, we will be publishing a series of tutorials on how to use the `ggplot2` package to create beautiful and informative data visualisations. Each tutorial will explain how to create a different type of plot, and will take you step-by-step from a basic plot to a highly customised graph.

We will demonstrate some of the many options the `ggplot2` package has for creating and customising line plots. We will use an international trade [dataset](http://pachamaltese.github.io/stats/trade-chile-china/copper-data-for-tutorial.csv ) (parts 1-4) made by ourselves from different sources (Chile Customs, Central Bank of Chile and General Directorate of International Economic Relations).

# Line Plots
```{r child = '1_Line_Plots.Rmd', cache=TRUE}
```

# Area Plots
```{r child = '2_Area_Plots.Rmd', cache=TRUE}
```

# Bar Plots
```{r child = '3_Bar_Plots.Rmd', cache=TRUE}
```

# Stacked Bar Plots
```{r child = '4_Stacked_Bar.Rmd', cache=TRUE}
```

# Scatter Plots
```{r child = '5_Scatter_Plots.Rmd', cache=TRUE}
```

# Weighted Scatter Plots
```{r child = '6_Weighted_Scatterplots.Rmd', cache=TRUE}
```

# Histograms
```{r child = '7_Histograms.Rmd', cache=TRUE}
```

# Density Plots
```{r child = '8_Density_Plots.Rmd', cache=TRUE}
```

# Function Plots
```{r child = '9_Function_Plots.Rmd', cache=TRUE}
```

# Boxplots
```{r child = '10_Boxplots.Rmd', cache=TRUE}
```

# Linear regression
```{r child = '11_Linear_Regression_Plot.Rmd', cache=TRUE}
```