---
title: "Creating plots in R using ggplot2 - Parts 1-9"
author: 
  - Jodie Burchell
  - Mauricio Vargas Sepúlveda 帕夏
date: '`r Sys.Date()`'
output:
  html_document:
    toc: yes
---

```{r setup, cache = FALSE, echo = FALSE, message = FALSE, warning = FALSE, tidy = FALSE}
knitr::opts_chunk$set(message = F, error = F, warning = F, comment = NA, fig.align = 'center', dpi = 100, fig.width=6, fig.height=5, tidy = F, cache.path = '.cache/', fig.path = '0_all_posts/')
```

I recently teamed up with [Mauricio Vargas Sepúlveda](http://pachamaltese.github.io/) to create some graphing tutorials in R. In the coming weeks, we will be publishing a series of tutorials on how to use the `ggplot2` package to create beautiful and informative data visualisations. Each tutorial will explain how to create a different type of plot, and will take you step-by-step from a basic plot to a highly customised graph.

We will demonstrate some of the many options the `ggplot2` package has for creating and customising line plots. We will use an international trade [dataset](http://pachamaltese.github.io/stats/trade-chile-china/copper-data-for-tutorial.csv ) (parts 1-4) made by ourselves from different sources (Chile Customs, Central Bank of Chile and General Directorate of International Economic Relations).

```{r child = '1_Line_Plots.Rmd'}
```

```{r child = '2_Area_Plots.Rmd'}
```

```{r child = '3_Bar_Plots.Rmd'}
```

```{r child = '4_Stacked_Bar.Rmd'}
```

```{r child = '5_Scatter_Plots.Rmd'}
```

```{r child = '6_Weighted_Scatterplots.Rmd'}
```

```{r child = '7_Histograms.Rmd'}
```

```{r child = '8_Density_Plots.Rmd'}
```

```{r child = '9_Function_Plots.Rmd'}
```